---
title: "Seasonal Prediction of Wildfires in the Apalachicola National Forest (ANF): Exploratory Analysis"
output: html_document
editor_options: 
  chunk_output_type: console
---

R version 4.1.2 (2021-11-01) -- "Bird Hippie"

## Get a boundary file for the ANF

https://data.fs.usda.gov/geodata/edw/edw_resources/shp/S_USA.NFSLandUnit.zip
```{r}
if(!"S_USA.NFSLandUnit" %in% list.files(here::here("data"))){
  download.file(url = "https://data.fs.usda.gov/geodata/edw/edw_resources/shp/S_USA.NFSLandUnit.zip", destfile = here::here("data", "S_USA.NFSLandUnit.zip"))
unzip(here::here("data", "S_USA.NFSLandUnit.zip"), exdir = here::here("data", "S_USA.NFSLandUnit"))
}

ANF_Boundary.sf <- sf::st_read(dsn = here::here("data", "S_USA.NFSLandUnit")) |>
  dplyr::filter(NFSLANDU_2 == "Apalachicola National Forest")

sf::st_area(ANF_Boundary.sf) # 2,567 sq. km (~634,000 acre)
```

Make a map showing the ANF boundary and the airport location. Export > Save as Image > tiff increasing width & height then open in Preview and convert to pdf.
```{r}
Airport.sf <- data.frame(Name = c("Regional Airport"), 
                         Latitude = c(30.39306),
                         Longitude = c(-84.35333)) |>
  sf::st_as_sf(coords = c("Longitude", "Latitude"),
               crs = 4326) |>
  sf::st_transform(crs = sf::st_crs(ANF_Boundary.sf))

tmap::tmap_mode("view")
tmap::tm_shape(ANF_Boundary.sf) +
  tmap::tm_borders() +
tmap::tm_shape(Airport.sf) +
  tmap::tm_dots(size = .03, 
                col = "darkgreen")
```

## Wildfire data and analysis

Severity of fires: Area burned, fire size 
Frequency of fires: Number of fires

```{r}
t0 <- Sys.time()
Fires.sf <- sf::st_read(dsn = here::here("data", "FPA_FOD_20210617.gpkg"),
                        layer = "Fires") |>
  dplyr::filter(STATE == "FL") |>
  sf::st_transform(crs = sf::st_crs(ANF_Boundary.sf)) |>
  sf::st_intersection(ANF_Boundary.sf)
Sys.time() - t0  # ~ 4 min on desktop computer, ~ 2.5 min on laptop
```

Add columns
```{r}
Fires.sf <- Fires.sf |>
  dplyr::mutate(Size = 
                dplyr::case_when(FIRE_SIZE_CLASS == "A" ~ 1,
                                 FIRE_SIZE_CLASS == "B" ~ 2,
                                 FIRE_SIZE_CLASS == "C" ~ 3,
                                 FIRE_SIZE_CLASS == "D" ~ 4,
                                 FIRE_SIZE_CLASS == "E" ~ 5,
                                 FIRE_SIZE_CLASS == "F" ~ 6,
                                 FIRE_SIZE_CLASS == "G" ~ 7)) |>
  dplyr::mutate(Year = lubridate::year(DISCOVERY_DATE),
                Month = lubridate::month(DISCOVERY_DATE,
                                         label = TRUE, 
                                         abbr = TRUE),
                DoY = lubridate::yday(DISCOVERY_DATE))
```

Bar chart of fire frequency by cause
```{r}
table(Fires.sf$NWCG_GENERAL_CAUSE) |>
  xtable::xtable()

df <- Fires.sf |>
  sf::st_drop_geometry() |>
  dplyr::group_by(NWCG_GENERAL_CAUSE) |>
  dplyr::summarize(nF = dplyr::n(),
                   perF = nF/nrow(Fires.sf)) |>
  xtable::xtable()

library(ggplot2)
ggplot(data = df,
       mapping = aes(y = reorder(NWCG_GENERAL_CAUSE, perF),
                     x = perF,
                     fill = perF)) +
  geom_col() +
  scale_fill_distiller(palette = "Oranges",
                       direction = 1,
                       guide = "none") +
  scale_x_continuous(labels = scales::percent) +
  ylab("") + xlab("") +
  labs(title = "Fires by cause in the Apalachicola National Forest",
       subtitle = "Based on data from 1992-2018",
       caption = "Data source: Short, Karen (2021)") +
  theme_minimal()
```

Bar chart of fire frequency by size class
```{r}
table(Fires.sf$Size) |>
  xtable::xtable()

df <- Fires.sf |>
  sf::st_drop_geometry() |>
  dplyr::filter(NWCG_CAUSE_CLASSIFICATION == "Natural") |>
  dplyr::group_by(Size) |>
  dplyr::summarize(nF = dplyr::n(),
                   perF = nF/nrow(Fires.sf)) |>
  xtable::xtable()

ggplot(data = df,
       mapping = aes(x = as.factor(Size),
                     y = perF,
                     fill = perF)) +
  geom_col() +
  scale_fill_distiller(palette = "Oranges",
                       direction = 1,
                       guide = "none") +
  scale_y_continuous(labels = scales::percent) +
  ylab("") + xlab("") +
  labs(title = "Fires by size class in the Apalachicola National Forest",
       subtitle = "Based on data from 1992-2018",
       caption = "Data source: Short, Karen (2021)") +
  theme_minimal()
```

Map showing the locations of all fires over the period of record.
```{r}
tmap::tmap_mode("plot")
tmap::tm_shape(ANF_Boundary.sf) +
  tmap::tm_borders() +
tmap::tm_shape(Fires.sf) +
  tmap::tm_bubbles(size = "Size", col = "orange") +
tmap::tm_shape(Airport.sf) +
  tmap::tm_dots(size = .1, col = "black") 
```

Monthly occurrence (relative frequency) of wildfires in the ANF.
```{r}
df <- Fires.sf |>
  sf::st_drop_geometry() |>
  dplyr::group_by(Month, .drop = FALSE) |>
  dplyr::summarize(nF = dplyr::n(),
                   perF = nF/nrow(Fires.sf))

ggplot(data = df,
      mapping = aes(x = Month, 
                    y = perF,
                    fill = perF)) +
  geom_col() +
  scale_fill_distiller(palette = "Oranges",
                       direction = 1,
                       guide = "none") +
  scale_y_continuous(labels = scales::percent_format(1L)) +
  labs(x = "", y = "Wildfire frequency") +
  theme_minimal() 
```

Graphs showing the annual seasonal number of fires and annual seasonal number of acres burned
```{r}
df <- Fires.sf |>
  sf::st_drop_geometry() |>
#  dplyr::filter(NWCG_CAUSE_CLASSIFICATION == "Natural") |>
  dplyr::filter(Month %in% c("May", "Jun", "Jul")) |>
  dplyr::mutate(YearF = factor(Year, 
                               levels = as.character(1992:2018), 
                               ordered = TRUE)) |>
  dplyr::group_by(YearF, .drop = FALSE) |>
  dplyr::summarize(nFires = dplyr::n(),
                   totalAcresBurned = sum(FIRE_SIZE),
                   AvgSizeCategory = mean(Size)) |>
  dplyr::mutate(Year = 1992:2018)

p1 <- ggplot(data = df,
             mapping = aes(x = Year, 
                           y = nFires)) +
  geom_point() +
  scale_x_continuous(breaks = seq(1992, 2018, by = 2)) +
  theme_minimal() +
  labs(x = "", y = "Number of wildfires")

p2 <- ggplot(data = df,
             mapping = aes(x = Year, 
                           y = totalAcresBurned)) +
  geom_point() +
  scale_x_continuous(breaks = seq(1992, 2018, by = 2)) +
  scale_y_log10(limits = c(1, 30000)) +
  theme_minimal() +
  labs(x = "", y = "Total acres burned")

p3 <- ggplot(data = df,
             mapping = aes(x = Year, 
                           y = AvgSizeCategory)) +
  geom_point() +
  scale_x_continuous(breaks = seq(1992, 2018, by = 2)) +
  theme_minimal() +
  labs(x = "", y = "Average size category")


library(patchwork)
p1 / p3
```

Annual statistics
```{r}
range(df$nFires)
df$Year[df$nFires == max(df$nFires)]
mean(df$nFires); var(df$nFires)
```

Seasonal temporal autocorrelation in frequency and total area burned.
```{r}
plot(acf(df$nFires))
plot(acf(df$totalAcresBurned))
cor.test(df$nFires, df$totalAcresBurned)
cor.test(df$nFires, df$AvgSizeCategory)
cor.test(df$nFires[-1], df$totalAcresBurned[-24])
```

No significant inter-seasonal autocorrelation. Significant correlation between number of fires and burned area as expected.

Make fire size an ordered factor and create a model that predicts the probability by size category similar to what we did for predicting tornado EF rating. See `PredictTorFreqByEF.Rmd` in the `EF-dist` distribution.
```{r}
Fires.sf |>
  sf::st_drop_geometry() |>
  dplyr::group_by(Size) |>
  dplyr::summarize(nFires = dplyr::n(),
                   Range1 = range(FIRE_SIZE)[1],
                   Range2 = range(FIRE_SIZE)[2])
```

## TLH WSO weather data

Import the summary of the day weather data and add columns to the data frame.
```{r}
TLH.df <- readr::read_csv(file = here::here("data", "TLH_Daily1940-2020.csv")) |>
  dplyr::mutate(Date = as.Date(DATE)) |>
  dplyr::mutate(Year = lubridate::year(Date), 
                Month = lubridate::month(Date, 
                                         label = TRUE, 
                                         abbr = TRUE),
                DoY = lubridate::yday(Date),
                MaxTemp = TMAX,
                MinTemp = TMIN,
                MaxTempC = (TMAX - 32) / 1.8,
                MinTempC = (TMIN - 32) / 1.8,
                Rainfall24 = PRCP,
                Rainfall24mm = Rainfall24 * 25.4) |>
  dplyr::filter(Year <= 2020)

sum(is.na(TLH.df$PRCP))
sum(is.na(TLH.df$PRCP)) / nrow(TLH.df) * 100  # only 5 missing values [~ .017% of all days]
sum(is.na(TLH.df$TMAX))
sum(is.na(TLH.df$TMAX)) / nrow(TLH.df) * 100  # only 1 missing value [~ .0034% of all days]

# Missing temperature and rainfall filled in from NCDC daily summaries https://gis.ncdc.noaa.gov/maps/ncei/summaries/daily

TLH.df$MaxTemp[TLH.df$Date == "2005-07-08"] <- 95
TLH.df$Rainfall24[is.na(TLH.df$PRCP)] <- c(.1, .3, 0, .1, 0)
TLH.df$Rainfall24mm[is.na(TLH.df$PRCP)] <- c(3, 8.9, 0, 2.5, 0)
```

Annual rainfall statistics
```{r}
AnnualRainfall <- TLH.df |>
  dplyr::filter(Year >= 1943) |>  # complete data started in March 1943
  dplyr::group_by(Year) |>
  dplyr::summarize(totalRainfall = sum(Rainfall24),
                   totalRainfallcm = totalRainfall * 2.54)

AnnualRainfall |> 
  dplyr::summarize(avgRainfall = mean(totalRainfallcm),
                   sdRainfall = sd(totalRainfallcm),
                   varRainfall = var(totalRainfallcm))
AnnualRainfall |>
  dplyr::arrange(totalRainfallcm)
```

## Compute KBDI

Compute daily values of the KBDI. Original paper outlining the rationale and how to create it: https://www.srs.fs.usda.gov/pubs/rp/rp_se038.pdf with a minor correction identified in Alexander1992.pdf (paper found in Dropbox/Literature).

Step one: Compute net rainfall on each day. Units are inches.
```{r}
source(here::here("scripts", "netRainfall.R"))
TLH.df$NetR <- netRainfall(Rainfall24 = TLH.df$Rainfall24)
```

Step two: Compute daily drought index. Units of temperature are degrees F. Q is set initially to 400. The influence of this start value diminishes to zero after a few months R is the annual average rainfall for TLH in inches
```{r}
source(here::here("scripts", "droughtIndex.R"))
DI <- droughtIndex(Q = 400, R = 60, 
                   MaxTemp = TLH.df$MaxTemp, 
                   NetR = TLH.df$NetR)
TLH.df$Ql <- DI$Ql
TLH.df$Qlm <- TLH.df$Ql * .254  # tenth of an inch to mm
TLH.df$DroughtIndex <- DI$DroughtIndex
```

Daily
```{r}
TLH.df <- TLH.df |>
  dplyr::filter(Year >= 1943) |>
  dplyr::mutate(Date2 = as.Date(doy - 1, origin = "2020-01-01"))

ggplot(data = TLH.df, 
       mapping = aes(x = Date2, y = Year, fill = Qlm)) +
  geom_tile() +
  scale_y_reverse(breaks = seq(1945, 2020, by = 10)) +
  scale_x_date(date_breaks = "month", 
               date_labels = "%b", 
               position = "top") +
  scale_fill_distiller(palette = "BrBG",
                       direction = -1,
                       name = "mm") +
  labs(x = "", y = "") +
  theme_minimal() 
```

Monthly average soil moisture deficit.
```{r}
TLH.df |>
  dplyr::mutate(Month = factor(month.abb[month], 
                               levels = month.abb, 
                               ordered = TRUE)) |>
  dplyr::group_by(Month, .drop = FALSE) |>
  dplyr::summarize(AvgSoilMoistureDeficit = mean(Qlm)) |>
ggplot(mapping = aes(x = Month, 
                     y = AvgSoilMoistureDeficit,
                     fill = AvgSoilMoistureDeficit)) +
  geom_col() +
  scale_fill_distiller(palette = "BrBG",
                       direction = -1,
                       guide = 'none') +
  labs(y = "Soil moisture deficit (mm)", x = "") +
  theme_minimal() 
```

Make some trend plots. This plot is statistically more realistic because it removes the temporal autocorrelations.
```{r}
TLH.df |>
  dplyr::group_by(month, Year) |>
  dplyr::summarise(MaxQlm = dplyr::last(Qlm)) |>
  dplyr::filter(month == "Apr") |>
ggplot(mapping = aes(x = Year, y = MaxQlm)) +
  geom_point() +
  geom_smooth(method = lm) +
  ylab("Soil moisture deficit (mm)") + xlab("") +
  theme_minimal() 
```

## Combine seasonal fire data and KBDI

```{r}
All.df <- Fires.sf |>
  sf::st_drop_geometry() |>
  dplyr::filter(Month %in% c("May", "Jun", "Jul")) |>
  dplyr::mutate(YearF = factor(Year, 
                               levels = as.character(1992:2018), 
                               ordered = TRUE)) |>
  dplyr::group_by(YearF, .drop = FALSE) |>
  dplyr::summarize(nFires = dplyr::n(),
                   TAB = sum(FIRE_SIZE))

Human.df <- Fires.sf |>
  sf::st_drop_geometry() |>
  dplyr::filter(NWCG_CAUSE_CLASSIFICATION == "Human") |>
  dplyr::filter(Month %in% c("May", "Jun", "Jul")) |>
  dplyr::mutate(YearF = factor(Year, 
                               levels = as.character(1992:2018), 
                               ordered = TRUE)) |>
  dplyr::group_by(YearF, .drop = FALSE) |>
  dplyr::summarize(nFires = dplyr::n(),
                   TAB = sum(FIRE_SIZE))

Natural.df <- Fires.sf |>
  sf::st_drop_geometry() |>
  dplyr::filter(NWCG_CAUSE_CLASSIFICATION == "Natural") |>
  dplyr::filter(Month %in% c("May", "Jun", "Jul")) |>
  dplyr::mutate(YearF = factor(Year, 
                               levels = as.character(1992:2018), 
                               ordered = TRUE)) |>
  dplyr::group_by(YearF, .drop = FALSE) |>
  dplyr::summarize(nFires = dplyr::n(),
                   TAB = sum(FIRE_SIZE)) 

SeasonalFires.df <- data.frame(Year = rep(1992:2018, times = 3),
                               Cause = rep(c("All", "Human", "Natural"), each = 27),
                               nFires = c(All.df$nFires, Human.df$nFires, Natural.df$nFires),
                               TAB = c(All.df$TAB, Human.df$TAB, Natural.df$TAB))
                               

Dryness.df <- TLH.df |>
  dplyr::filter(Year >= 1992 & Year <= 2018) |>
  dplyr::filter(month == "Apr") |>
  dplyr::group_by(Year) |>
  dplyr::summarize(QlmLastDay = dplyr::last(Qlm),
                   AvgQlm = mean(Qlm))

SeasonalFires.df <- SeasonalFires.df |>
  dplyr::left_join(Dryness.df, by = "Year")

readr::write_csv(SeasonalFires.df, 
                 file = here::here("data", "SeasonalFires.csv"))
```

Correlations and scatter plots
```{r}
SeasonalFires.df |>
  dplyr::group_by(Cause) |>
  dplyr::summarize(rQlmFires = cor(QlmLastDay, nFires),
                   rQlmTAB = cor(QlmLastDay, TAB))

SeasonalFires.df |>
ggplot(mapping = aes(x = QlmLastDay, 
                     y = nFires, 
                     label = Year)) +
  geom_point() +
 # ggrepel::geom_label_repel() +
  facet_wrap(~ Cause) +
  geom_smooth(span = .95, se = FALSE) +
  labs(x = "Soil moisture deficit (mm) on April 30th", 
       y = "Number of fires in the ANF (May-Jul)") +
  theme_minimal()

SeasonalFires.df |> 
  dplyr::mutate(logTAB = log(TAB + .1)) |>
ggplot(mapping = aes(x = QlmLastDay, 
                     y = logTAB, 
                     label = Year)) +
  geom_point() +
 # ggrepel::geom_label_repel() +
 # scale_y_log10() +
  facet_wrap(~ Cause) +
  geom_smooth(method = lm, se = FALSE) +
  labs(x = "Soil moisture deficit (mm) on April 30th", 
       y = "Log of total acres burned in the ANF (May-Jul)") +
  theme_minimal()
```

Correlation by day of month during April
```{r}
df <- SeasonalFires.df
TLH.df |>
  dplyr::filter(Year >= 1992 & Year <= 2018) |>
  dplyr::filter(month == "Apr") |>
  dplyr::group_by(doy, Year) |>
  dplyr::mutate(Qlm = Qlm) |>
  dplyr::select(Year, doy, Qlm) |>
  dplyr::filter(doy != 91 & doy != 121) |>
  dplyr::group_by(doy) |>
  dplyr::summarize(r = cor.test(Qlm, df$nFires)$estimate,
            rLo = cor.test(Qlm, df$nFires, conf.level = .9)$conf.int[1],
            rHi = cor.test(Qlm, df$nFires, conf.level = .9)$conf.int[2],
            rs = cor(Qlm, df$nFires, method = "s")) |>
  xtable::xtable()
```

Make a plot
```{r}
  ggplot(data = df, 
         mapping = aes(x = Date, y = r)) +
  geom_ribbon(aes(ymin = rLo, ymax = rHi), fill = "gray90") +
  geom_point() +
  geom_line() +
  scale_y_continuous(limits = c(0, 1)) +
#  scale_x_continuous(breaks = c(1,	32,	60,	91, 121, 152, 182, 213, 244, 274, 305, 335),
#                     labels = month.abb, position = "top") +
  theme_minimal()
```

A slight upward trend in the correlation from early to late April but no significant difference.

Continue with `FireCountModel.Rmd`.

Test models
```{r}
df <- SeasonalFires.df |>
  dplyr::filter(Cause == "Human")

xtable::xtable(MASS::glm.nb(nFires ~ QlmLastDay, data = df))

summary(lm(I(log(totalAcresBurned + .01)) ~ QlmLastDay, data = SeasonalFires.df)) 
```


Statistical significance of the trend by month.
```{r}
summary(lm(Qlm ~ Date, data = TLH.df)) # this model assumes daily Qlm are independent (they are certainly not)

library(broom)

( Trends.df <- TLH.df %>% 
  group_by(Year, month) %>%
  summarize(AvgQlm = mean(Qlm)) %>%
  group_by(month) %>%
  do(tidy(lm(AvgQlm ~ Year, data = .))) %>%
  dplyr::filter(term == "Year") )

Avg <- TLH.df %>%
  group_by(month) %>%
  summarize(AvgQlm = mean(Qlm)) %>%
  pull(AvgQlm)

cor(Trends.df$estimate, Avg) # positive, indicating upward trends are occurring in months with higher drought index values

Avg <- TLH.df %>%
  group_by(month) %>%
  summarize(nYear = n_distinct(Year), 
            TotalPRCP = sum(PRCP, na.rm = TRUE),
            AvgMonthlyPRCP = TotalPRCP/nYear) %>%
  pull(AvgMonthlyPRCP)

cor(Trends.df$estimate, Avg)  # negative as expected since upward trends are occurring in months with less rainfall
```

Soil moisture deficits are increasing at .3 mm/year on average during April. Which amounts to 3 mm/decade.

What is the correlation between PRCP, TMAX and Qlm grouped by month?
```{r}
TLH.df %>%
  group_by(month) %>%
  summarize(rPRCP = cor(Qlm, PRCP, use = "complete"),
            rTMAX = cor(Qlm, TMAX, use = "complete"),
            rTMIN = cor(Qlm, TMIN, use = "complete"),
            rTAVG = cor(Qlm, TAVG, use = "complete"))
```

Seasonal (fire season) temperature vs KDBI scatter plot. Create a character variable called `FireSeason` (May, June, July). Then group by that variable computing the average temp and total precipitation (or avg KDBI).
```{r}
df <- TLH.df %>%
  mutate(FireSeason = month %in% c("May", "Jun", "Jul")) %>%
  filter(FireSeason) %>%
  group_by(Year) %>%
  summarize(AvgHiTemp = mean(MaxTemp),
            AvgLoTemp = mean(MinTemp),
            TotalPrcp = sum(Rainfall24mm),
            TotalNetR = sum(NetR),
            AvgQlm = mean(Qlm))

ggplot(df, 
       mapping = aes(y = TotalNetR, x = AvgHiTemp, label = Year, color = factor(Year))) +
  geom_point() +
  geom_label_repel(size = 2.5) +
  scale_y_continuous(limits = c(0, NA)) +
  scale_color_grey(start = .8, end = .2, guide = "none") +
#  geom_smooth(method = lm, se = FALSE, col = "black") +
  ylab("Total Precipitation (mm)") + xlab("Average Daily High Temperature (°F)") +
  ggtitle(label = "Fire seasons in the Apalachicola National Forest (May-Jul) that are drier tend to be hotter",
          subtitle = "Dry days are sunnier and thus they tend to get hotter. Later years have darker labels.") +
  theme_minimal() 
```

SOI is positively correlated with moisture deficit. La Nina conditions (positive SOI) during winter into spring leads to larger moisture deficits (higher Qlm). The upward trend in Qlm remains significant after accounting for SOI.

In terms of frequency by Drought Index category. Proportion of the year in each drought category.
```{r}
TLH.df %>%
  group_by(Year, DroughtIndex) %>%
  summarize(N = n()) %>%
  ggplot(aes(x = Year, y = N)) +
  geom_col(aes(fill = as.factor(DroughtIndex))) +
  scale_fill_manual(values = terrain.colors(8), name = "KDBI") +
  scale_x_continuous(name = "", breaks = seq(1941, 2020, 4)) +
  scale_y_reverse() +
  theme_minimal() +
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  guides(fill = guide_legend(reverse = TRUE)) +
  labs(title = "The proportion of the year in drought (KDBI > 5) in the ANF has increased since 1960",
       subtitle = "KDBI = 7-6: Drought, 4-5: Dry, 1-3: Moist, 0: Saturated",
       caption = "Period of record: 1941-2020, Data source: NWSFO Tallahassee") 
```

Number of days per year with KBDI above 650 (hundredths of an inch) (Qlm = 165.1)
```{r}
library(MASS)

TLH.df |>
  group_by(Year) |>
  summarize(nD = sum(Qlm > 160)) |>
  ggplot(aes(Year, nD)) +
  geom_point() +
  labs(x = "", y = "Number of days\nSoil moisture deficit exceeded 160 mm") +
#  labs(title = "Number of days KBDI exceeds 160 mm",
#       x = "", y = "") +
#  stat_smooth(method = "glm.nb",
#              formula = y ~ x, 
#              se = FALSE,
#              col = "red") +
  theme_minimal()
```

