---
title: "Seasonal prediction of natural wildfires in the Apalachicola National Forest (ANF)"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Get the required packages
```{r}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(MASS)
library(brms)
library(tidybayes)
library(modelr)
library(equatiomatic)
library(xtable)
```

## Get the data

The data are organized in the file `ANF_Fires.Rmd`. Change moisture deficit units to cm.
```{r}
SeasonalData.df <- read_csv(file = "SeasonalForecastModelData.csv") %>%
  mutate(QlmLastDay = QlmLastDay / 10) %>% # change units from mm to cm
  mutate(nFiresDiff = nFiresAll - nFires) %>%
  mutate(AcresBurnedDiff = AcresBurnedAll - AcresBurned)

W <- SeasonalData.df$nFiresAll
```

```{r}
var(W)/mean(W) #13.6

model0 <- glm.nb(W ~ 1)

summary(model0)

model1 <- glm.nb(nFiresAll ~ QlmLastDay,
                 data = SeasonalData.df)

summary(model1)

xtable(model1)
extract_eq(model1, intercept = "beta")
```
$$
\log ({ E( \operatorname{nFires} ) })  = \beta_{0} + \beta_{1}(\operatorname{QlmLastDay})
$$

In-sample skill estimates
```{r}
What <- predict(model1, type = "response")
cor(W, What)
mean(abs(W - What))
mean((W - What)^2)
```

Hold-one-out cross validation (out of sample skill estimates)
```{r}
What <- vector()
for(i in 1:dim(SeasonalData.df)[1]){
  model <- glm.nb(nFires ~ QlmLastDay, data = SeasonalData.df[-i,])
  What[i] <- predict(model, type = "response", 
                    newdata = data.frame(QlmLastDay = SeasonalData.df$QlmLastDay[i]))
}
cor(W, What)
mean(abs(W - What))
mean((W - What)^2)
```

Graph showing observed values and predicted line.
```{r}
ggplot(data = SeasonalData.df, 
       mapping = aes(x = QlmLastDay, y = nFires)) +
  geom_point() + 
  geom_line(aes(y = fitted(model1))) +
  ylab("Number of Fires (May-July)") + xlab("Soil Moisture Deficit (April)") +
  theme_minimal()

model2 <- glm.nb(nFires ~ QlmLastDay + preYearAcresBurned,
                 data = SeasonalData.df)
coef(model2) |>
  xtable()
```

Use the {brms} and {tidybayes} packages to get posterior distributions.

Also hot-days.Rmd, PredictTorFreqByEF.Rmd

When using {brms} cite the following publications:

Bürkner P. C. (2017). brms: An R Package for Bayesian Multilevel Models using Stan. Journal of Statistical Software. 80(1), 1-28. doi.org/10.18637/jss.v080.i01

Bürkner P. C. (2018). Advanced Bayesian Multilevel Modeling with the R Package brms. The R Journal. 10(1), 395-411. doi.org/10.32614/RJ-2018-017

As {brms} is a high-level interface to Stan, please additionally cite Stan:

Carpenter B., Gelman A., Hoffman M. D., Lee D., Goodrich B., Betancourt M., Brubaker M., Guo J., Li P., and Riddell A. (2017). Stan: A probabilistic programming language. Journal of Statistical Software. 76(1). 10.18637/jss.v076.i01

```{r}
fit0 <- brm(nFires ~ 1,
            family = "negbinomial",
            data = SeasonalData.df)

prior_summary(fit0)

stancode(fit0)

plot(fit0, variable = "b_Intercept")
```

Note `b_Intercept` is the mean (mu) and `shape` is the size parameter when generating random numbers with the `rnbinom()` function.

Posterior samples of the parameters (intercept and shape).
```{r}
ps <- fit0 |>
  posterior::as_draws_df()
```

Posterior samples of the outcome (number of fires).
```{r}
pp <- fit0 |>
  brms::posterior_predict() |>
  as.data.frame()
```

Each column is a posterior predictive sample for a particular outcome value.

## Visualizing uncertain

See https://www.youtube.com/watch?v=wbzfqh_3LyM https://mjskay.github.io/ggdist/articles/slabinterval.html

Posterior samples of the parameters using {tidybayes}
```{r}
library(tidybayes)

get_variables(fit0)

fit0 |>
  spread_draws(b_Intercept, shape) %>%
  head(10)

fit0 |>
  spread_draws(b_Intercept) |>
  ggplot(mapping = aes(x = b_Intercept)) +
#    stat_halfeye()
#    stat_gradientinterval()
    stat_dotsinterval()
```

Posterior samples of the _expected_ outcome for each outcome value using {tidybayes}
```{r}
ps <- SeasonalData.df |>
  add_epred_draws(fit0)

SeasonalData.df |>
  add_epred_draws(fit0) |>
  ggplot(mapping = aes(x = .epred)) +
  facet_wrap(~ Year) +
  stat_dotsinterval() +
  geom_point(data = SeasonalData.df,
             mapping = aes(y = 0, x = nFires),
             color = "red")
```
  
This shows the uncertainty of the expected count of the number fires. It does not contain the uncertainty associated with a particular count given the expected count.

```{r}
pp <- fit0 |>
  brms::posterior_predict() |>
  as.data.frame()

colnames(pp) <- seq(1992, 2018)

pp |>
  pivot_longer(cols = "1992":"2018") |>
  rename(Year = name,
         Count = value)

pp |>
  pivot_longer(cols = "1992":"2018") |>
  rename(Year = name,
         Count = value) |>
  ggplot(mapping = aes(x = Count)) +
  facet_wrap(~ Year) +
  stat_dotsinterval() +
  geom_point(data = SeasonalData.df,
             mapping = aes(y = 0, x = nFires),
             color = "red")
```

This shows the uncertainty of the expected counted _together_ with the uncertainty associated with a particular count given the expected count.

Create a new model with dryness as a predictor.
```{r}
fit1 <- brm(nFires ~ QlmLastDay,
            family = "negbinomial",
            data = SeasonalData.df)

prior_summary(fit1)
stancode(fit1)
```

Posterior draws of the coefficient on dryness.
```{r}
fit1 |>
  spread_draws(b_QlmLastDay) |>
  head(15)

fit1 |>
  spread_draws(b_QlmLastDay) |>
  ggplot(mapping = aes(x = b_QlmLastDay)) +
    stat_dotsinterval()
```

```{r}
ps <- SeasonalData.df |>
  tidyr::expand(QlmLastDay = modelr::seq_range(QlmLastDay, n = 25)) |> #1D GRID
  tidybayes::add_epred_draws(fit1)
 
ps <- SeasonalData.df |>
  tidyr::expand(QlmLastDay = modelr::seq_range(QlmLastDay, n = 25)) |> #GRID
  tidybayes::add_epred_draws(fit1) |> #CONDITION
  tidybayes::median_qi(.epred, .width = c(.66, .95)) #MUNGE

ggplot() +
  geom_lineribbon(mapping = aes(x = QlmLastDay,
                     y = .epred,
                     ymin = .lower,
                     ymax = .upper),
                  data = ps) +
  scale_fill_brewer() +
  geom_point(mapping = aes(x = QlmLastDay, y = nFires),
             color = "orange",
             data = SeasonalData.df) +
  theme_minimal()

SeasonalData.df |>
  tidyr::expand(QlmLastDay = modelr::seq_range(QlmLastDay, n = 25)) |> #GRID
  tidybayes::add_epred_draws(fit1) |> #CONDITION
  dplyr::filter(.draw <= 20) |>
  ggplot() +
  geom_line(mapping = aes(x = QlmLastDay, y = .epred, group = .draw), 
            color = "skyblue") +
  geom_point(mapping = aes(x = QlmLastDay, y = nFires),
             color = "orange",
             data = SeasonalData.df) +
  labs(x = "Soil Moisture Deficit on April 30th (cm)",
       y = "Number of Fires in the ANF (May-July)") +
  theme_minimal()
```

Total acres burned
```{r}
ggplot(data = SeasonalData.df,
       mapping = aes(x = AcresBurned)) +
  geom_density() +
  geom_rug() 

fit2 <- brm(I(AcresBurned + .01) ~ QlmLastDay,
            family = "exponential",
            data = SeasonalData.df)

prior_summary(fit2)

plot(fit2, pars = "QlmLastDay")

fit2 |>
  spread_draws(b_QlmLastDay) |>
  ggplot(mapping = aes(b_QlmLastDay)) +
    stat_dotsinterval()

ps <- SeasonalData.df |>
  tidyr::expand(QlmLastDay = modelr::seq_range(QlmLastDay, n = 25)) |> #GRID
  tidybayes::add_epred_draws(fit2) |> #CONDITION
  tidybayes::median_qi(.epred, .width = c(.66, .95)) #MUNGE

ggplot() +
  geom_lineribbon(mapping = aes(x = QlmLastDay,
                     y = .epred,
                     ymin = .lower,
                     ymax = .upper),
                  data = ps) +
  scale_fill_brewer() +
  geom_point(mapping = aes(x = QlmLastDay, y = AcresBurned),
             color = "orange",
             data = SeasonalData.df) +
  theme_minimal()

SeasonalData.df |>
  tidyr::expand(QlmLastDay = modelr::seq_range(QlmLastDay, n = 25)) |> #GRID
  tidybayes::add_epred_draws(fit2) |> #CONDITION
  dplyr::filter(.draw <= 20) |>
  ggplot() +
  geom_line(mapping = aes(x = QlmLastDay, y = .epred, group = .draw), 
            color = "skyblue") +
  geom_point(mapping = aes(x = QlmLastDay, y = AcresBurned),
             color = "orange",
             data = SeasonalData.df) +
  theme_minimal()
```

When more than one predictor variable, try the {marginaleffects} package http://arelbundock.com/brms.html to get posterior draws with user-specified values and other predictors held at their means.